Probabilistic Catalogs for Crowded Stellar Fields

### 2013-03-10 paper/response.md

Brendon J. Brewer
Daniel Foreman-Mackey
David W. Hogg

Itemized Responses
==================

"Abstract seems to be overselling the case in multiple instances; for example, "introduce" may be too strong given Stetson's work on DAOPHOT"

- We have considerably reworked the abstract to more moderate phrasing. Related approaches definitely exist
and we have added a few more citations that we have found out about since (e.g.
Hobson 2003).

@@

"- p. 2: PHOTO should be photo"

- Fixed

"p. 2, first paragraph: Can't you simplify this long discussion by just stating that catalogs are lists of best-fit model parameters?"

- Sometimes they are lists of best fitting model parameters (by some criterion of "best"), but sometimes they are not. It depends on how they were produced. SExtractor, for example, doesn't really resembling fitting. It's a point estimation procedure.
We have significantly restructured the introduction to try to make this and following discussion clearer, and to separate out the philosophical idea of doing things the Bayesian way from the practical effects of what can happen in crowded fields.

@@

"very lossy", "destroy information" etc. - this makes it sound like the end of the world. Either be quantitative or add appropriate caveats; (overselling this method may be counterproductive).

- In principle it is of course lossy and destroys information, but the degree to which this is practically important is not clear yet. We have moderated our language to this effect.

"p. 2, third paragraph: I am *very* confused by the statement that "Instead of estimating..." First, there is nothing more than "estimating" that we can do whatever the method. Second, and more importantly, standard methods (as in SExtractor, photo, and most other codes) *do* estimate errors for flux, positions, and many other parameters, and these error estimates *are* propagated (in most papers) to subsequent analysis. What is *not* probabilistic about giving flux+-error?"

- This is a language issue. We were using estimation in the statistics sense of "point estimates obtained from some procedure". Standard errors are related to probability but there are many ways of obtaining them in the community. We have tried to be consistent about saying point estimation where that is what we mean.

@@

"general comment on Introduction: typically, one knows what exactly the paper is about after reading it - not the case here."

- Hopefully our changes to the introduction make this more clear. In particular we have added a paragraph describing what we discuss in each following section.

@@

"minor comment: eqs. 14 and 15: don't you want the tolerance to be +-0.1*(xmax-xmin)? That is, what if xmin=0?"

- This was an error, it has been fixed

"Section 5: words like BIC and Bayesian model comparison are never mentioned; looking at the left panel in Fig. 5 I wonder if high-N models should be penalized for having more parameters (or if perhaps you already included this effect); I also wonder if the scatter in h1 and other global parameters is much larger than for the case when all the fluxes are perfectly known and only LF is determined."

- BIC is an approximation that shouldn't be used (in my opinion). Model comparison is implicit in this analysis and doesn't need to be added on after the fact, we have added a short discussion about this to the Test Case 1 results section. That is because N is a parameter that we can propose to change. The Occam penalty occurs naturally in our model without us having to do anything.

@@

"fig. 4: pretty but useless! What is the reader supposed to do - stare at it and count all the points? It would be better to use images such as Fig. 1 and 2, and show the mean image for all catalogs, and then a few residual images (perhaps normalized by statistical errors)."

- We do show these summary images in Figure 5. We would like to retain Figure 4 primarily because it shows the fundamental output of this approach, which is samples, not point estimates. We also like the fact that it is pretty!

"fig. 5: what are the stretch and scale in the lower left panel? How do residuals compare to statistical fluctuations?"

- The residuals are just the statistical fluctuations and pass all traditional tests such as reduced chi-squared. This essentially has to be true because we know the model is correct. In practice this may not be the case on real data, in which
case the model assumptions (PSF etc) would need to be modified. This is discussed in the paragraph that begins "From these samples".

@@

"Section 6: Fig. 6 is very misleading (it implies that SExtractor, which was so many times successfully tested, is lousy). A much more informative plot (I tried it in practice in a similar context) would be to plot cumulative counts vs. log(flux) (bright to faint) for your method and for SExtractor on the same panel. Additional useful plot would be to match sources from the two lists within some small radius and compare their photometry and astrometry with true values. It wouldn't be surprising if SExtractor has biases in crowded field case, and such biases would be *a really strong* argument in favor of your method!"

- We have replaced Figure 6 with the suggested luminosity function plot. It shows that the SExtractor catalogs lie within the uncertainties from the posterior distribution but *only* at the bright end, as you suspect. Our method outperforms SExtractor at the faint end. This can be partially explained by our use of a known parametric form for the LF. To clarify this we have added some discussion of this point.

"Section 6: running SExtractor down to 0.5sigma and computing LF with all these sources is pretty silly (as the sample will have a lot of faint junk; also large errors and Malmquist bias will be important). I find the clustering of alpha1 ~ alpha2 ~ 1 strongly suspect. Also, even for a clean sample with vanishing errors, there are many pitfalls when fitting power-law LFs. I strongly suggest the authors take a look at Clauset, Shalizi, and Newman (2009). Power-Law Distributions in Empirical Data. SIAM Review 51, 661-703."

- I do not understand how else one would compute the LF from a traditional catalog apart from computing the catalog and then fitting the power-law. We are aware of the pitfalls of power laws and have added a caveat saying that it's mostly for demonstration purposes. However, we have removed this figure because it is too dramatic. I do not understand what the referee means by "I find the clustering of alpha1 ~ alpha2 ~ 1 strongly suspect". We believe that there are no major technical errors in our work.

### 2013-03-10 paper/ms.tex
Changed a sentence
Use xrange in definition of pixel width
Changes to revisions after Hogg email

Throughout this paper we will
assume the pixel-convolved PSF is a weighted mixture of two concentric
circular
Gaussians (a similar mixture model that uses a Gaussian core and flexible wings is also used by DAOPHOT, \citet{1987PASP...99..191S}) with widths
$s_1$ and $s_2$:
\begin{eqnarray}
\mathcal{P}(x, y) = \frac{w}{2\pi s_1^2}\exp
\left[
-\frac{1}{2s_1^2}\left(x^2 + y^2\right)
\right]
+ \frac{1-w}{2\pi s_2^2}\exp
\left[-\frac{1}{2s_2^2}\left(x^2 + y^2\right)
\right].
\end{eqnarray}
The pixellated observed image is assumed to be generated from the mock
image (evaluated at the center of each pixel) plus noise:
\begin{eqnarray}
I_{ij} &=& \mathcal{M}(X_{ij}, Y_{ij}) + \epsilon_{ij}
\end{eqnarray}
where the errors $\{\epsilon_{ij}\}$ are independent and normally distributed.
The variance of the normal distribution for each pixel is determined by the
brightness of the sky and the
brightness of the mock image in that pixel. This can be modelled by assuming the
following distribution:
\begin{eqnarray}
\epsilon_{ij} \sim \mathcal{N}(0, \sigma_0^2 + \eta \mathcal{M}(X_{ij}, Y_{ij})).
\end{eqnarray}
where $\sigma_0$ is a constant noise level and $\eta$ is a coefficient allowing
the noise to be higher in brighter regions of the image. This dependence of the
noise variances on the model intensity arises as a result of the Poissonian
nature of photon counts, but allows for the fact that a ``sky'' background may
have already been subtracted from the image in the reduction process.
This parameterisation has been used by \citet{2011MNRAS.412.2521B} and is an
alternative to the common practice of producing a ``variance map'' from the
image data that is then assumed to be known.

@@

The priors for the PSF parameters and the noise parameters were assigned to be:
\begin{eqnarray}
\ln s_1 &\sim& \textnormal{Uniform}(\ln(0.3L), \ln(30L)) \\
\ln s_2 &\sim& \textnormal{Uniform}(\ln(s_1), \ln(s_1) + 2.3) \\
w &\sim& \textnormal{Uniform}(0, 1)\\
\ln \sigma_0 &\sim& \textnormal{Uniform}(\ln(10^{-3}), \ln(10^{3})) \\
\ln \eta &\sim& \textnormal{Uniform}(\ln(10^{-3}), \ln(10^{3}))
\end{eqnarray}
where $L = (x_{\rm max} - x_{\rm min})/n$ is the width of a pixel. These priors
describe vague prior knowledge about the overall scale of the PSF except that the
wider component is less than 10 times as wide as the narrow component, as well
as the knowledge that the noise variance is not extreme relative to the fluxes
of the stars.

@@

\subsection{Test Case 1}
Test Case 1 was run with the DNS algorithm and usable results were obtained
within about an hour on a modern desktop PC. The inferences on the parameters
$N$, $h_1$, $h_2$, $\alpha_1$, and $\alpha_2$ are shown in
Figure~\ref{fig:results1}. The number of stars is correctly inferred, and the
posterior distributions for the other parameters comfortably contain the true
input values. Note that the uncertainty in $h_2$, $\alpha_1$, and $\alpha_2$ is
quite large. This is because the broken power-law model
(Figure~\ref{fig:powerlaw}) does not change drastically in shape as the
parameters are varied. Therefore, a large number of stars would be required to
tightly constrain these parameters.

@@

This result may be attributable to the fact that the Bayesian method knows
about the existence and form of the luminosity function, even though it
does not know the values of the parameters. To test this, we ran the inference
on the data using an incorrect exponential distribution for the luminosity
function. The resulting CLF from this run did underestimate the number of low stars.
However, the
Bayesian evidence for the exponential model was significantly lower (by a factor of
approximately $10^7$) than
for the (correct) broken power-law model.

@@

In this paper we have developed and demonstrated a Bayesian approach to
making catalogs from astronomical images in the case where the image contains
only stars (or other point sources). The key idea is that instead of computing
a single catalog, the method creates a posterior probability distribution on
the space of possible catalogs that represents our state of knowledge about
the presence and properties of objects in the image. When this is done, the
uncertainties in the imaging are accurately propagated through to scientific
conclusions, for example about the luminosity function of the stars.
This approach was contrasted with the results from the standard \sex~software.
For the bright sources the results were essentially consistent, however the
Bayesian approach was more successful at modelling the distribution of faint
stars. Of course, the Bayesian method is much more computationally intensive,
which is a significant issue in practice. To build a more complete picture of
when our approach is necessary in practice, it will need to be tested against
a wider variety of alternative methods  such as DAOPHOT \citep{1987PASP...99..191S}
and DOLPHOT \citep{dolphot}.

>>>>

Throughout this paper we will
assume the pixel-convolved PSF is a weighted mixture of two concentric
circular
Gaussians (a similar mixture model that uses a Gaussian core and flexible wings is also used by DAOPHOT, \citet{1987PASP...99..191S}) with widths
$s_1$ and $s_2$:
\begin{eqnarray}
\mathcal{P}(x, y) = \frac{w}{2\pi s_1^2}\exp
\left[
-\frac{1}{2s_1^2}\left(x^2 + y^2\right)
\right]
+ \frac{1-w}{2\pi s_2^2}\exp
\left[-\frac{1}{2s_2^2}\left(x^2 + y^2\right)
\right].
\end{eqnarray}
The pixellated observed image is assumed to be generated from the mock
image (evaluated at the center of each pixel) plus noise:
\begin{eqnarray}
I_{ij} &=& \mathcal{M}(X_{ij}, Y_{ij}) + \epsilon_{ij}
\end{eqnarray}
where the errors $\{\epsilon_{ij}\}$ are independent and normally distributed.
The variance of the normal distribution for each pixel is determined by the
brightness of the sky and the
brightness of the mock image in that pixel. This can be modelled by assuming the
following distribution:
\begin{eqnarray}
\epsilon_{ij} \sim \mathcal{N}(0, \sigma_0^2 + \eta \mathcal{M}(X_{ij}, Y_{ij})).
\end{eqnarray}
where $\sigma_0$ is a constant noise level and $\eta$ is an unknown = coefficient that allows for the possibility that
the noise level is higher in brighter regions of the image. This dependence of the
noise variances on the model intensity arises as a result of the Poissonian
nature of photon counts, but allows for the fact that a ``sky'' background may
have already been subtracted from the image in the reduction process.
This parameterisation has been used by \citet{2011MNRAS.412.2521B} and is an
alternative to the common practice of producing a ``variance map'' from the
image data that is then assumed to be known.

@@

The priors for the PSF parameters and the noise parameters were assigned to be:
\begin{eqnarray}
\ln s_1 &\sim& \textnormal{Uniform}(\ln(0.3L), \ln(30L)) \\
\ln s_2 &\sim& \textnormal{Uniform}(\ln(s_1), \ln(s_1) + 2.3) \\
w &\sim& \textnormal{Uniform}(0, 1)\\
\ln \sigma_0 &\sim& \textnormal{Uniform}(\ln(10^{-3}), \ln(10^{3})) \\
\ln \eta &\sim& \textnormal{Uniform}(\ln(10^{-3}), \ln(10^{3}))
\end{eqnarray}
where $L = x_{\rm range}/n$ is the width of a pixel. These priors
describe vague prior knowledge about the overall scale of the PSF except that the
wider component is less than 10 times as wide as the narrow component, as well
as the knowledge that the noise variance is not extreme relative to the fluxes
of the stars.

@@

\subsection{Test Case 1}
Test Case 1 was run with the DNS algorithm and usable results were obtained
within about an hour on a modern desktop PC. The inferences on the parameters
$N$, $h_1$, $h_2$, $\alpha_1$, and $\alpha_2$ are shown in
Figure~\ref{fig:results1}. The number of stars is correctly inferred, and the
posterior distributions for the other parameters comfortably contain the true
input values. As $N$ is a parameter of our model, there is no need for Bayesian
``Model Comparison'' calculations to be done between different values of $N$.
The DNS method does compute the ``evidence'' value that is required for model
comparison, but this is useful only to test completely separate models, it
is not needed to infer the value of $N$.

Note that the uncertainty in $h_2$, $\alpha_1$, and $\alpha_2$ is
quite large. This is because the broken power-law model
(Figure~\ref{fig:powerlaw}) does not change drastically in shape as the
parameters are varied. Therefore, a large number of stars would be required to
tightly constrain these parameters.

@@

This result may be attributable to the fact that the Bayesian method knows
about the existence and form of the luminosity function, even though it
does not know the values of the parameters. To test this, we ran the inference
on the data using an incorrect exponential distribution for the luminosity
function. The resulting CLF from this run did undershoot the true CLF
at the faint end.
However, the
Bayesian evidence for the exponential model was significantly lower (by a factor of
approximately $10^7$) than
for the (correct) broken power-law model.

In practice, we note that the wings of the PSF might become degenerate
with an nonzero flat background level in the data. To test whether this
was influencing the inferences (particularly about the faint end of the CLF)
we also ran a model that included an unknown constant background. This
had only a minor effect on the resulting inferences.

@@

In this paper we have developed and demonstrated a Bayesian approach to
making catalogs from astronomical images in the case where the image contains
only stars (or other point sources). The key idea is that instead of computing
a single catalog, the method creates a posterior probability distribution on
the space of possible catalogs that represents our state of knowledge about
the presence and properties of objects in the image. When this is done, the
uncertainties in the imaging are accurately propagated through to scientific
conclusions, for example about the luminosity function of the stars.
This approach was contrasted with the results from the standard \sex~software.
For the bright sources the results were essentially consistent, however the
Bayesian approach was more successful at modelling the distribution of faint
stars. Of course, the Bayesian method is much more computationally intensive,
which is a significant issue in practice.
However, the great value of upcoming imaging data sets, and the irreproducibility of astronomical imaging data in areas of time-domain astrophysics (for example in observations of rare events), make it important to extract as much information as possible from every patch of imaging.  Our view is that the additional CPU time, and the non-triviality of the outputs from our method, will be worth the effort in the next generation of astronomical experiments.
To build a more complete picture of
when our approach is necessary in practice, it will need to be tested against
a wider variety of alternative methods  such as DAOPHOT \citep{1987PASP...99..191S}
and DOLPHOT \citep{dolphot}.
